---
title: "How to Run Fiddle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Run Fiddle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

FIDDLE stands for *F*lex*I*ble *D*ata-*D*riven Preprocessing Pipe*L*in*E* for Structured Clinical Data. This acronym may be a bit of a stretch, but it is a proposed methodology to standardize the data-preprocessing used in data analysis requiring machine learning. You can read more about FIDDLE in [JAMIA](https://doi.org/10.1093/jamia/ocaa139) or the [FIDDLE GitHub page](https://github.com/MLD3/FIDDLE).

As FIDDLE was originally written in Python, this package is meant to provide an R wrapper to help those unfamiliar with Python to implement this pipeline in R. This vignette will show you how to run FIDDLE using the `run_fiddle` function.


# Setup

To begin, first load this package, `fiddleR`. Next, you need to setup a python environment which has the necessary libraries to run FIDDLE. You can do this by running the following code:

```{r setup, eval = FALSE}
library(fiddleR)

# Setup python environment
setup_python_env(envname = "test_env")
```

This code will create a new python environment in the current working directory called `test_env`. This environment will have the necessary libraries to run FIDDLE.


# Running FIDDLE

FIDDLE requires a few inputs, namely

1. `data`: The path to the data file you intend to use for your ML analysis.
2. `population`: The path to the a file that contains a vector of the population you intend to use for your ML analysis.
3. `T_`:  The time of prediction; time-dependent features will be generated using data in $t \in [0, T]$.
4. `dt`: The temporal granularity for which you want to look at time dependent data.
5. `output_dir`: The directory where you want to save the output files.
6. `config_file`: The path to the configuration file you want to use.

Here is an example of a call to FIDDLE:


```{r, eval = FALSE}
dat <- system.file("extdata", "test1", "data.csv", package = "fiddleR")
pop <- system.file("extdata", "test1",  "pop.csv", package = "fiddleR")
config_path <- system.file("extdata", "test1", "config-1-parallel.yaml", package = "fiddleR")
output_dir <- tempdir()


run_fiddle(
  data = dat, 
  population = pop, 
  T_ = 4, 
  dt = 1, 
  output_dir = output_dir,
  config_file = config_path,
  theta_1 = 0.001, 
  theta_2 = 0.001, 
  theta_freq = 1
)
```

Note that when you run FIDDLE, the output directory will include some new files. Namely:

- `s.npz`: a sparse array of shape (N, d)
- `X.npz`: a sparse tensor of shape (N, L, D)
- `s.feature_names.json`: names of _d_ time-invariant features
- `X.feature_names.json`: names of _D_ time-series features
- `x.feature_aliases.json`: aliases of duplicated time-invariant features
- `X.feature_aliases.json`: aliases of duplicated time-series features

You will end up using these files in your machine learning analysis. To load these into an R format, you can use the `load_data` function here.

```{r, eval=FALSE}
tensors = load_data(output_dir)

# Access the data
s = tensors$s
x = tensors$x
```
